---
output: html_document
editor_options: 
  chunk_output_type: inline
---

# Explaining the Predictive Performance of Black Box Classification Models
## *Luis Torgo, Paulo Azevedo, Ines Areosa*

This page contains the code, data and resulting plots of the experiments in the paper: *Explaining the Predictive Performance of Black Box Classification Models*.

First, it is provided an explanation on how the full experiments can be reproduced, followed with an example of the EDP tool. 

Finally we provide a zip file with plots corresponding to all 18 data sets, not included in the paper due to space limitation.


## Requirements
[R](www.r-project.org) is required for reproducing the experiments. The following R extra packages, available on [CRAN](https://cran.r-project.org/web/packages/), are also needed:

* dplyr
* GGally
* plyr
* gridExtra
* reshape2


## Usage

Tested on R 4.0.2

The full code and the data sets can be downloaded [here](code.zip). The resultant plots can be downloaded [here](plots.zip)


### Load Functions and Dataset

```{r, eval=FALSE,warning=FALSE}
## Loading functions required (as well as required packages)
## source("performance_functions.R")

## Loading Dataset
load("dataSetsWithPreds.Rdata")
names(DSsPreds) <- c('a1','a2','a3','a4','a6','a7','Abalone','acceleration','availPwr','bank8FM','cpuSm','fuelCons','boston','maxTorque','machineCpu','servo','airfoild','concreteStrength')
nmod <- 4

```

### Run the Experiments for the Full Benchmark Dataset

```Rscript DataSet_Performance.R```

### An Example on How to Employ EDP Functions

Choose what type of error the end user wants to analyse: "Absolute", Logarithmic ("Log") or "Residual"
```{r eval=FALSE}
err <- "Absolute"
```

Choose a particular data set from the full benchmark
```{r eval=FALSE}
y_observed <- "servo"
dataset <- DSsPreds[[y_observed]][complete.cases(DSsPreds[[y_observed]]),]
#Show part of dataset
head(dataset, n=10)
```

Choose which model is to be analysed and prepare the data
```{r eval=FALSE}
mod <- "svm"
single_error_ds <- single_model_data(data=dataset[1:(length(dataset) - nmod)], model=dataset[[mod]], feature_y=names(dataset)[1], type=err)
head(single_error_ds, n=10)
```


Choose a predictor and calculate the EDP. Here is the example with a numeric ("vgain") and a categorical ("screw"), as well as the bivariate EDP for both features:

```{r, eval=FALSE, echo=FALSE, results='hide', fig.keep='all'}
edp(single_error_ds, "vgain", type=err, jitter=FALSE)
edp(single_error_ds, "screw", type=err, jitter=FALSE)

# Bivariate EDP: Option "All" for the comparative facet to show theoverall error distribution
plot(multiple_edp(single_error_ds, c("vgain", "screw"), type=err, option='All', ncols=3, mode="wrap"))
```









