---
output: html_document
editor_options: 
  chunk_output_type: inline
---

# Explaining the Predictive Performance of Black Box Classification Models
## *Luis Torgo, Paulo Azevedo, Ines Areosa*

This page contains the code, data and resulting plots of the experiments in the paper: *Explaining the Predictive Performance of Black Box Classification Models*.

## Experiments of the Univariate Analysis (based on EDPs)

First, we provide an explanation on how the full experiments can be reproduced, followed with an example of the EDP tool. 

Finally we provide a zip file with plots corresponding to all 18 data sets, not included in the paper due to space limitation.


### Software Requirements
[R](www.r-project.org) is required for reproducing the experiments. The following R extra packages, available on [CRAN](https://cran.r-project.org/web/packages/), are also needed:


* dplyr
* gridExtra
* reshape2
* RColorBrewer
* gtable
* ggplot2

Tested on R 4.0.2

### Usage

The full code and the data sets can be downloaded [here](code.zip). The resultant plots can be downloaded [here](plots.zip)


#### Load Functions and Dataset

```{r, eval=FALSE, warning=FALSE}
## Loading functions required (as well as required packages)
source('Functions_ClassificationEDP.R')


## Loading Dataset
load("benchmarkWithPreds.Rdata")

```

#### Run the Experiments for the Full Benchmark Dataset

```Rscript Calculate_results.R```

### An Example on How to Employ EDP Function


Choose a particular data set from the full benchmark. Here an example for "Adult". Column *Class* shows the true prediction, while *svm*, *randomForest*, *nnet* and *naiveBayes* show the predictions for each of the models, as described in the paper
```{r eval=FALSE}

data_set <- DSOut[["Adult"]][complete.cases(DSOut[["Adult"]]),]

#Show part of dataset
head(data_set, n=10)
```

Choose which model is to be analysed and prepare the data. The last field, *details*, defines the information to present the error:

- "show_all" shows what was the model prediction versus the true prediction (when wrong)
- "reduced" shows only what category did the model fail to predict
```{r eval=FALSE}
mod <- "svm"

db <- prep_data(data_set[,-which(names(data_set) %in% c("svm","randomForest", "nnet", "naiveBayes"))], data_set[, mod], "Class", details = "show_all") #First field should include entire dataset excepting model predictions; Second field should include the model predictions; third field should be the name of the predicted class
    
head(db, n=10)
```


Choose a predictor and calculate the EDP. Here is the example with a numeric ("capital.gain") and a categorical ("relationship").
There is an option for additional zoom in the error part, that can be added by including the condition, *zoom = TRUE*, as the example in the second EDP.
The numerical feature bins can be controlled one of the three features:

- nquant: Number of bins (of equal size)
- size: size of bins
- quantil_break: Breaks in terms of quantiles
- breaks: List of breaking points

```{r, eval=FALSE, results='hide', fig.keep='all'}
cep(db, "capital.gain")

cep(db, "relationship", zoom = TRUE)
```

## Experiments of the Multivariat Analysis (based on Caren)

Caren is linux command-line java-based software to derive association rules. It also implements distribution rules, a form of rules to study patterns of a numeric property of interest.
In this paper we developed a categorical version of these rules to shed light on abnormal behavior in the performance of classification models. You can download this version [here](caren-2.6.5.zip)


### Data

The dataset formated used in this study is (example with breast-base dataset):

```
age,menopause,tumor.size,inv.nodes,node.caps,deg.malig,breast,breast.quad,irradiat,CM
'40-49','premeno','30-34','0-2','no','3','left','left_up','no',0
'60-69','ge40','30-34','0-2','no','3','left','left_low','no',0
'40-49','premeno','25-29','0-2','no','1','right','right_low','no',0
'60-69','ge40','40-44','3-5','yes','3','right','left_low','no',21
...
```

where CM is the categorical property of interest. In this example it represent the confusion matrix obtained from model SVM in dataset breast-base. '0' represents an hit
and '21' codifies an error where an example of class '2' was predicted with class '1'.


### comand line commands

Check caren switch with command 
```
> java -cp . caren -help
```

Example of used command to derive rules for dataset breast-base

```
> java -cp . caren d:\PJA\caren\TorgoPescas\ines\breast-base_svm.csv  0.01  0.5  -Att  -s,  -GCM  -Cat  -ovrb1  -Dist  -imp0.00000000000001  -null?  -fr
```

All obtained rules were derived using similar commands i.e. same minsup (0.01), same pruning filter (-imp), same nulls treatment (-null?), relative frequency of bins in distribution (-fr), etc.


#### Example of a rule:

```
Ant sup = 0.03750  pvalue = 0.0010004171846829683000  
CM={ 12/0.333,21/0.667 }    <--    breast.quad='right_up'  &  menopause='ge40'
```


