---
title: Predicting Algae Blooms
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---


```{r echo=FALSE}
knitr::opts_chunk$set(eval =FALSE)
```


The following is a [script file](RCode/Ralgae.R) containing all R code of all sections in this chapter.


<!-- loadData.Rnw --> 


### Loading the Data into R
```{r}
data(algae, package="DMwR2")
algae
```


```{r}
algae <- read.table('Analysis.txt',
           header=FALSE,
           dec='.',
           col.names=c('season','size','speed','mxPH','mnO2','Cl',
           'NO3','NH4','oPO4','PO4','Chla','a1','a2','a3','a4',
           'a5','a6','a7'),
           na.strings=c('XXXXXXX'))
```


```{r}
tibble::as_tibble(algae)
```


<!-- vizSum.Rnw --> 
```{r}
opts_template$set(onlyShow=list(echo=TRUE, eval=FALSE,  tidy=FALSE),
                  onlyRun=list(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE),
                  showFig=list(fig.width=6,fig.height=6,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,out.width="0.7\\textwidth"),
                  showFig2=list(fig.width=12,fig.height=6,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,out.width="0.9\\textwidth"),
                  runShow=list(echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, tidy=FALSE))
library(ggplot2)
library(grid)
library(DMwR2)
library(dplyr)
data(algae)
```




### Data Visualization and Summarization
```{r}
data(algae, package="DMwR2")
summary(algae)
```


```{r}
library(ggplot2)
ggplot(algae,aes(x=mxPH)) + geom_histogram(aes(y=..density..))
```


```{r}
library(ggplot2)
ggplot(algae,aes(x=mxPH)) + geom_histogram(aes(y=..density..))
```


```{r}
ggplot(algae,aes(x=mxPH)) + 
    geom_histogram(aes(y=..density..)) + 
    geom_density(color="red") + geom_rug() + 
    ggtitle("The Histogram of mxPH (maximum pH)") + 
    xlab("") + ylab("")
library(car)
qqPlot(algae$mxPH,main='Normal QQ plot of maximum pH',ylab="")
```


```{r}
library(car)
gh <- ggplot(algae,aes(x=mxPH)) + geom_histogram(aes(y=..density..)) + geom_density(color="red") + geom_rug() + ggtitle("The Histogram of mxPH (maximum pH)") + xlab("") + ylab("")
par(mfrow=c(1,2))
vpL <- viewport(height=unit(1, "npc"), width=unit(0.5, "npc"), 
                           just="left", 
                           y=0.5, x=0)
qqPlot(algae$mxPH,main='Normal QQ plot of maximum pH')
print(gh,vp=vpL)
qqPlot(algae$mxPH,main='Normal QQ plot of maximum pH',ylab="")
```


```{r}
ggplot(algae,aes(x=factor(0),y=oPO4)) + 
    geom_boxplot() + geom_rug() + 
    geom_hline(aes(yintercept=mean(algae$oPO4, na.rm = TRUE)),
               linetype=2,colour="red") +
    ylab("Orthophosphate (oPO4)") + xlab("") + scale_x_discrete(breaks=NULL)
```


```{r}
ggplot(algae,aes(x=factor(0),y=oPO4)) + 
    geom_boxplot() + geom_rug() + 
    geom_hline(aes(yintercept=mean(algae$oPO4, na.rm = TRUE)),linetype=2,colour="red") +
    ylab("Orthophosphate (oPO4)") + xlab("") + scale_x_discrete(breaks=NULL)
```


```{r}
plot(algae$NH4, xlab = "")
abline(h = mean(algae$NH4, na.rm = T), lty = 1)
abline(h = mean(algae$NH4, na.rm = T) + sd(algae$NH4, na.rm = T), lty = 2)
abline(h = median(algae$NH4, na.rm = T), lty = 3)
identify(algae$NH4)
```


```{r}
plot(algae$NH4, xlab = "")
clickedRows <- identify(algae$NH4)
algae[clickedRows, ]
```


```{r}
library(dplyr)
filter(algae, NH4 > 19000)
```


```{r}
ggplot(algae,aes(x=size,y=a1)) + geom_boxplot() +
    xlab("River Size") + ylab("Algal A1")
```


```{r}
library(forcats)
algae <- mutate(algae,
                size=fct_relevel(size,c("small","medium","large")),
                speed=fct_relevel(speed,c("low","medium","high")),
                season=fct_relevel(season,c("spring","summer","autumn","winter")))
```


```{r}
ggplot(algae,aes(x=size,y=a1)) + geom_boxplot() +
    xlab("River Size") + ylab("Algal A1")
```


```{r}
ggplot(algae,aes(x=size,y=a1)) + 
    geom_violin() + geom_jitter() + xlab("River Size") + ylab("Algal A1")
```


```{r}
ggplot(algae,aes(x=size,y=a1)) + geom_violin() + geom_jitter() + xlab("River Size") + ylab("Algal A1")
```


```{r}
data2graph <- filter(algae,!is.na(mnO2)) %>%
    mutate(minO2=cut(mnO2, quantile(mnO2,c(0,0.25,.5,.75,1)), include.lowest=TRUE))
ggplot(data2graph,aes(x=a3,y=season, color=season)) + geom_point() + 
    facet_wrap(~ minO2) + 
    guides(color=FALSE)
```


```{r}
data2graph <- filter(algae,!is.na(mnO2)) %>%
    mutate(minO2=cut(mnO2, 
                     quantile(mnO2,c(0,0.25,0.5,0.75,1)), 
                     include.lowest=TRUE))
#data2graph <- algae[!is.na(algae$mnO2),]
#data2graph <- cbind(data2graph,
#                    minO2=cut(data2graph$mnO2,
#                              quantile(data2graph$mnO2,c(0,0.25,.5,.75,1)),
#                              include.lowest=TRUE))
ggplot(data2graph,aes(x=a3,y=season,col=season)) + geom_point() + facet_wrap(~ minO2) + guides(color=FALSE)
```


<!-- Unknown.Rnw --> 
```{r}
opts_template$set(onlyShow=list(echo=TRUE, eval=FALSE,  tidy=FALSE),
                  onlyRun=list(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE),
                  showFig=list(fig.width=6,fig.height=6,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,out.width="0.7\\textwidth"),
                  showFig2=list(fig.width=12,fig.height=6,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,out.width="0.9\\textwidth"),
                  runShow=list(echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, tidy=FALSE))
library(ggplot2)
library(grid)
library(DMwR2)
library(dplyr) 
library(forcats)
data(algae)
algae <- mutate(algae,
                size=fct_relevel(size,c("small","medium","large")),
                speed=fct_relevel(speed,c("low","medium","high")),
                season=fct_relevel(season,c("spring","summer","autumn","winter")))
```




### Unknown Values
```{r}
library(DMwR2)
library(dplyr)  
data(algae)  
```




#### Removing the Observations with Unknown Values
```{r}
filter(algae, !complete.cases(algae) )
```


```{r}
algae <- na.omit(algae)
```


```{r}
data(algae, package="DMwR2") # only necessary if you executed the above na.omit()
algae <- algae[-c(62, 199), ]
```


```{r}
apply(algae, 1, function(x) sum(is.na(x)))
```


```{r}
data(algae, package="DMwR2")
manyNAs(algae, 0.2)
```


```{r}
algae <- algae[-manyNAs(algae), ]
```




#### Filling in the Unknowns with the Most Frequent Values
```{r}
algae[48, "mxPH"] <- mean(algae$mxPH, na.rm = TRUE)
```


```{r}
algae[is.na(algae$Chla), "Chla"] <- median(algae$Chla, na.rm = TRUE)
```


```{r}
data(algae, package="DMwR2")
algae <- algae[-manyNAs(algae), ]
algae <- centralImputation(algae)
```




#### Filling in the Unknown Values by Exploring Correlations
```{r}
cor(algae[, 4:18], use = "complete.obs")
```


```{r}
symnum(cor(algae[,4:18],use="complete.obs"))
```


```{r}
library(corrplot)
cm <- cor(algae[,4:18], use="complete.obs")
corrplot(cm, type="upper", tl.pos="d")
corrplot(cm, add=TRUE, type="lower", method="number", 
         diag=FALSE, tl.pos="n", cl.pos="n")
```


```{r}
library(corrplot)
cm <- cor(algae[,4:18],use="complete.obs")
corrplot(cm,type="upper",tl.pos="d",tl.cex=0.75)
corrplot(cm,add=TRUE, type="lower", method="number",tl.cex=0.75, diag=FALSE,tl.pos="n", cl.pos="n")
```


```{r}
data(algae, package="DMwR2")
algae <- algae[-manyNAs(algae), ]
lm(PO4 ~ oPO4, data = algae)
```


```{r}
algae[28, "PO4"] <- 42.897 + 1.293 * algae[28, "oPO4"]
```


```{r}
data(algae, package="DMwR2")
algae <- algae[-manyNAs(algae), ]
fillPO4 <- function(oP) ifelse(is.na(oP),NA,42.897 + 1.293 * oP)
algae[is.na(algae$PO4), "PO4"] <- sapply(algae[is.na(algae$PO4), "oPO4"], fillPO4)
```


```{r}
library(ggplot2)
library(forcats)
algae <- mutate(algae,
                size=fct_relevel(size,c("small","medium","large")),
                speed=fct_relevel(speed,c("low","medium","high")),
                season=fct_relevel(season,c("spring","summer","autumn","winter")))
ggplot(algae, aes(x=mxPH)) + geom_histogram(binwidth=0.5) + facet_wrap(~ season)
```


```{r}
library(ggplot2) 
algae <- mutate(algae,
                size=fct_relevel(size,c("small","medium","large")),
                speed=fct_relevel(speed,c("low","medium","high")),
                season=fct_relevel(season,c("spring","summer","autumn","winter")))
ggplot(algae, aes(x=mxPH)) + geom_histogram(binwidth=0.5) + facet_wrap(~ season)
```


```{r}
ggplot(algae, aes(x=mxPH)) + geom_histogram(binwidth=0.5) + 
    facet_wrap(size ~ speed)
```


```{r}
ggplot(algae, aes(x=mxPH, y=size, color=size)) + geom_point() + 
    facet_wrap(~speed) + geom_jitter(height = 0.4)
```


```{r}
ggplot(algae, aes(x=mxPH, y=size, color=size)) + geom_point() + facet_wrap(~speed) + geom_jitter(height = 0.4)
```




#### Filling in the Unknown Values by Exploring Similarities between Cases
```{r}
data(algae, package="DMwR2")
algae <- algae[-manyNAs(algae), ]
```


```{r}
algae <- knnImputation(algae, k = 10)
```


```{r}
algae <- knnImputation(algae, k = 10, meth = "median")
```


<!-- LinRegr.Rnw --> 
```{r}
opts_template$set(onlyShow=list(echo=TRUE, eval=FALSE,  tidy=FALSE),
                  onlyRun=list(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE),
                  showFig=list(fig.width=6,fig.height=6,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE),
                  showFig2=list(fig.width=12,fig.height=6,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE),
                  runShow=list(echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, tidy=FALSE))
library(DMwR2)
library(dplyr)
```




#### Multiple Linear Regression
```{r}
data(algae, package="DMwR2")
algae <-  algae[-manyNAs(algae), ]
clean.algae <- knnImputation(algae, k = 10)
```


```{r}
lm.a1 <- lm(a1 ~ ., data = clean.algae[, 1:12])
```


```{r}
summary(lm.a1)
```


```{r}
anova(lm.a1)
```


```{r}
lm2.a1 <- update(lm.a1, . ~ . - season)
```


```{r}
summary(lm2.a1)
```


```{r}
anova(lm.a1,lm2.a1)
```


```{r}
final.lm <- step(lm.a1)
```


```{r}
summary(final.lm)
```


<!-- RTs.Rnw --> 
```{r}
opts_template$set(onlyShow=list(echo=TRUE, eval=FALSE,  tidy=FALSE),
                  onlyRun=list(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE),
                  showFig=list(fig.width=6,fig.height=6,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,out.width="0.7\\textwidth"),
                  showFig2=list(fig.width=12,fig.height=6,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,out.width="0.9\\textwidth"),
                  runShow=list(echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, tidy=FALSE))
library(DMwR2)
library(dplyr)
```




#### Regression Trees
```{r}
library(rpart)
data(algae, package="DMwR2") 
algae <- algae[-manyNAs(algae), ]
rt.a1 <- rpart(a1 ~ ., data = algae[, 1:12])
```


```{r}
rt.a1
```


```{r}
library(rpart.plot)
prp(rt.a1,extra=101,box.col="orange",split.box.col="grey")
```


```{r}
library(rpart.plot)
prp(rt.a1,extra=101,box.col="orange",split.box.col="grey")
```


```{r}
printcp(rt.a1)
```


```{r}
rt2.a1 <- prune(rt.a1, cp = 0.08)
rt2.a1
```


```{r}
(rt.a1 <- rpartXse(a1 ~ ., data = algae[, 1:12]))
```


```{r}
first.tree <- rpart(a1 ~ ., data = algae[, 1:12])
snip.rpart(first.tree, c(4, 7))
```


```{r}
plot(first.tree)
text(first.tree)
snip.rpart(first.tree)
```


<!-- EvalAndSel.Rnw --> 
```{r}
opts_template$set(onlyShow=list(echo=TRUE, eval=FALSE,  tidy=FALSE),
                  onlyRun=list(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE),
                  showFig=list(fig.width=6,fig.height=6,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,out.width="0.7\\textwidth"),
                  showFig2=list(fig.width=12,fig.height=6,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,out.width="0.9\\textwidth"),
                  runShow=list(echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, tidy=FALSE))
library(ggplot2)
library(DMwR2)
library(dplyr)
library(rpart)
data(algae)
algae <- algae[-manyNAs(algae), ]
clean.algae <- knnImputation(algae, k = 10)
lm.a1 <- lm(a1 ~ ., data = clean.algae[, 1:12])
final.lm <- step(lm.a1)
rt.a1 <- rpart(a1 ~ ., data = algae[, 1:12])
```




### Model Evaluation and Selection
```{r}
lm.predictions.a1 <- predict(final.lm, clean.algae)
rt.predictions.a1 <- predict(rt.a1, algae)
```


```{r}
(mae.a1.lm <- mean(abs(lm.predictions.a1 - algae[["a1"]])))
(mae.a1.rt <- mean(abs(rt.predictions.a1 - algae[["a1"]])))
```


```{r}
(mse.a1.lm <- mean((lm.predictions.a1 - algae[["a1"]])^2))
(mse.a1.rt <- mean((rt.predictions.a1 - algae[["a1"]])^2))
```


```{r}
(nmse.a1.lm <- mean((lm.predictions.a1-algae[['a1']])^2)/
               mean((mean(algae[['a1']])-algae[['a1']])^2))
(nmse.a1.rt <- mean((rt.predictions.a1-algae[['a1']])^2)/
               mean((mean(algae[['a1']])-algae[['a1']])^2))
```


```{r}
library(ggplot2)
dg <- data.frame(lm.a1=lm.predictions.a1,
                 rt.a1=rt.predictions.a1,
                 true.a1=algae[["a1"]])
ggplot(dg,aes(x=lm.a1,y=true.a1)) +
    geom_point() + geom_abline(slope=1,intercept=0,color="red") +
    ggtitle("Linear Model")
ggplot(dg,aes(x=rt.a1,y=true.a1)) +
    geom_point() + geom_abline(slope=1,intercept=0,color="red") +
    ggtitle("Regression Tree")
```


```{r}
library(grid)
dg <- data.frame(lm.a1=lm.predictions.a1,rt.a1=rt.predictions.a1,true.a1=algae[["a1"]])
g1 <- ggplot(dg,aes(x=lm.a1,y=true.a1)) +
    geom_point() + geom_abline(slope=1,intercept=0,color="red") +
    ggtitle("Linear Model")
g2 <- ggplot(dg,aes(x=rt.a1,y=true.a1)) +
    geom_point() + geom_abline(slope=1,intercept=0,color="red") +
    ggtitle("Regression Tree")
grid.newpage()
pushViewport(viewport(layout=grid.layout(1,2)))
print(g1,vp=viewport(layout.pos.row=1,layout.pos.col=1))
print(g2,vp=viewport(layout.pos.row=1,layout.pos.col=2))
```


```{r}
plot(lm.predictions.a1,algae[['a1']],main="Linear Model",
     xlab="Predictions",ylab="True Values")
abline(0,1,col="red")
algae[identify(lm.predictions.a1,algae[['a1']]),]
```


```{r}
sensible.lm.predictions.a1 <- ifelse(lm.predictions.a1 < 0, 0, lm.predictions.a1)
(mae.a1.lm <- mean(abs(lm.predictions.a1 - algae[["a1"]])))
(smae.a1.lm <- mean(abs(sensible.lm.predictions.a1 - algae[["a1"]])))
```


```{r}
library(performanceEstimation)
res <- performanceEstimation(
    PredTask(a1 ~ ., algae[, 1:12], "a1"),
    c(Workflow(learner="lm",pre="knnImp",post="onlyPos"),
      workflowVariants(learner="rpartXse",learner.pars=list(se=c(0,0.5,1)))),
    EstimationTask(metrics="nmse",method=CV(nReps=5,nFolds=10))
)
```


```{r}
summary(res) 
```


```{r}
plot(res)
```


```{r}
plot( res ) 
```


```{r}
getWorkflow("rpartXse.v1", res)
```


```{r}
DSs <- sapply(names(algae)[12:18],
          function(x,names.attrs) { 
            f <- as.formula(paste(x, "~ ."))
            PredTask(f, algae[,c(names.attrs,x)], x, copy=TRUE) 
          },
          names(algae)[1:11])
res.all <- performanceEstimation(
    DSs,
    c(Workflow(learner="lm", pre="knnImp", post="onlyPos"),
      workflowVariants(learner="rpartXse", learner.pars=list(se=c(0,0.5,1)))),
    EstimationTask(metrics="nmse" ,method=CV(nReps=5, nFolds=10)))
```


```{r}
plot(res.all)
```


```{r}
plot(res.all)
```


```{r}
topPerformers(res.all)
```


```{r}
library(randomForest)
res.all <- performanceEstimation(
    DSs,
    c(Workflow(learner="lm", pre="knnImp",post="onlyPos"),
      workflowVariants(learner="rpartXse",
                       learner.pars=list(se=c(0,0.5,1))),
      workflowVariants(learner="randomForest", pre="knnImp",
                       learner.pars=list(ntree=c(200,500,700)))),
    EstimationTask(metrics="nmse",method=CV(nReps=5,nFolds=10)))
```


```{r}
rankWorkflows(res.all, top=3)
```


```{r}
p <- pairedComparisons(res.all,baseline="randomForest.v3")
p$nmse$F.test
p$nmse$BonferroniDunn.test
```


```{r}
CDdiagram.BD(p)
```


<!-- Preds.Rnw --> 
```{r}
opts_template$set(onlyShow=list(echo=TRUE, eval=FALSE,  tidy=FALSE),
                  onlyRun=list(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE),
                  showFig=list(fig.width=6,fig.height=6,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,out.width="0.7\\textwidth"),
                  showFig2=list(fig.width=12,fig.height=6,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,out.width="0.9\\textwidth"),
                  runShow=list(echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, tidy=FALSE))
library(ggplot2)
library(DMwR2)
library(performanceEstimation)
library(randomForest)
data(algae, package="DMwR2")
algae <- algae[-manyNAs(algae), ]
load("algResAll.Rdata")
```




### Predictions for the Seven Algae
```{r}
wfs <- sapply(taskNames(res.all),
              function(t) topPerformer(res.all,metric="nmse",task=t))
wfs[["a1"]]
wfs[["a7"]]
```


```{r}
full.test.algae <- cbind(test.algae, algae.sols)
pts <- array(dim = c(140,7,2),
             dimnames = list(1:140, paste0("a",1:7), c("trues","preds")))
for(i in 1:7) {
    res <- runWorkflow(wfs[[i]],
                       as.formula(paste(names(wfs)[i],"~.")),
                       algae[,c(1:11,11+i)],
                       full.test.algae[,c(1:11,11+i)])
    pts[,i,"trues"] <- res$trues
    pts[,i,"preds"] <- res$preds
}
```


```{r}
pts[1:3,c("a1","a3"),]
```


```{r}
avg.preds <- apply(algae[,12:18], 2, mean)
apply((pts[,,"trues"] - pts[,,"preds"])^2, 2 ,sum) /
    apply( (scale(pts[,,"trues"], avg.preds, FALSE))^2, 2, sum)
```


